\mutagen: Coverage-Guided, Property-Based Testing using Exhaustive Type-Preserving Mutations

### 2021-08-24 papers/icse22/main.tex
More general compression
Compressing. Compressing.

\usepackage{wrapfig}
\usepackage{xcolor}
\usepackage{color}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{caption}
\usepackage[ruled,vlined,noend]{algorithm2e}
\usepackage{xspace}
\usepackage{tikz}
\usepackage{tikzscale}
\usepackage{pgfplots}
\usepackage{paralist}
\usepackage{enumitem}
%% \usepackage[square,numbers]{natbib}

@@

@@

\begin{verbatim}
gen_tree :: Size -> Gen a -> Gen (Tree a)
gen_tree 0 gen_a = Leaf <$> gen_a
gen_tree n gen_a = oneof
  [ Leaf <$> gen_a
  , Branch <$> gen_tree (n-1) <*> gen_a <*> gen_tree (n-1) ]
\end{verbatim}

\noindent This definition follows a common type-directed generation approach,
and it is a good example of what to expect from a random generator synthesized
automatically from the codebase.
%
In esence, this generator randomly picks one data constructor from its
corresponding data type definition with uniform probability (i.e, either a
\texttt{Leaf} or a \texttt{Branch}), and then calls itself for every recursive
subexpression needed to produce a well-typed value, carefully reducing the size
limit by a unit on each step, ensuring termination by restricting itself to
generating leaves when it reaches zero.
%
Payloads are generated simply by delegating the task to an external random
generator (\texttt{gen\_a}) defined elsewhere.
%

@@

Readers familiar with RBPT will recognize that attempting to test
\texttt{prop\_insert\_inv} using this generator in \quickcheck directly does not
work well.
%
In practice, \quickcheck will give up quite if most of the generated test cases
get discarded because they do not satisfy the balanced precondition.
%
Sadly, most of the inputs produced by our na\"ive generator suffer from this
problem, and the interesting part of the property (\texttt{balanced (insert x
  t)}) is tested very sporadically as a result.
%
In other words, \emph{the precondition of our testing property becomes very
  sparse when we use simple, automatically derived generators.}

@@

%% Once inside of the main loop, the algorithm picks the next test case using a
%% simple criterion: if there are valid values enqueued for mutation, it picks the
%% first one, mutates it and returns it, decreasing its energy by one.
%% %
%% If \textit{QValid} is empty, then the same is attempted using \textit{QDisc}.
%% %
%% If none of the mutation queues contain any candidates, CGPT generates a new
%% value from scratch.
%% %
%% This selection process is illustrated in detail in Algorithm
%% \ref{algo:fuzzchick:pick}.

@@

%% \paragraph{Testing Loop}
%% Outlined in Algorithm \ref{algo:fuzzchick:loop}, CGPT starts by creating two
%% queues, \textit{QValid} and \textit{QDisc} for valid and discarded previously
%% executed test cases, respectively.
%% %
%% Enqueued values are stored along with a given mutation budget, that controls how
%% many times a given test case can be mutated before being finally thrown away.

%% Once inside of the main loop, the algorithm picks the next test case using a
%% simple criterion: if there are valid values enqueued for mutation, it picks the
%% first one, mutates it and returns it, decreasing its energy by one.
%% %
%% If \textit{QValid} is empty, then the same is attempted using \textit{QDisc}.
%% %
%% If none of the mutation queues contain any candidates, CGPT generates a new
%% value from scratch.
%% %
%% This selection process is illustrated in detail in Algorithm
%% \ref{algo:fuzzchick:pick}.

@@

%% \begin{algorithm}
%%   \SetArgSty{textnormal}
%%   %% \SetInd{0em}{0.75em}
%%   \SetKw{KwNot}{not}
%%   \SetAlgoLined
%%   \DontPrintSemicolon
%%   \SetKwFunction{FC}{Loop}
%%   \SetKwProg{Fn}{Function}{:}{}
%%   \Fn{\FC{P, N, gen, mut}}{
%%     i $\gets$ 0\;
%%     TLog, QValid, QDisc $\gets$ $\varnothing$\;
%%     \While{i $<$ N}{
%%       x $\gets$ Pick(QValid, QDisc, gen, mut)\;
%%       (result, trace) $\gets$ WithTrace(P(x))\;
%%       \lIf{\KwNot result}{\KwRet Bug(x)}
%%       \If{Interesting(TLog, trace)}{
%%         e $\gets$ Energy(TLog, x, trace)\;
%%         \eIf{\KwNot Discarded(result)}{
%%           Enqueue(QValid, (x, e))\;
%%         }{
%%           Enqueue(QDisc, (x, e))\;
%%         }
%%       }
%%       i $\gets$ i+1\;
%%     }
%%     \KwRet Ok\;
%%   }
%% \caption{\label{algo:fuzzchick:loop}CGPT Testing Loop}
%% \end{algorithm}

%% \begin{algorithm}
%%   \SetArgSty{textnormal}
%%   %% \SetInd{0em}{0.75em}
%%   \SetKw{KwNot}{not}
%%   \SetAlgoLined
%%   \DontPrintSemicolon
%%   \SetKwFunction{Pick}{Pick}
%%   \SetKwProg{Fn}{Function}{:}{}
%%   \Fn{\Pick{QValid, QDisc, gen, mut}}{
%%     \If{\KwNot Empty(QValid)}{
%%       (x,e) $\gets$ Deque(QValid)\;
%%       \If{e $>$ 0}{
%%         PushFront(QValid, (x, e-1))
%%       }
%%       \KwRet Sample(mut(x))\;
%%     }
%%     \uElseIf{\KwNot Empty(QDisc)}{
%%       (x,e) $\gets$ Deque(QDisc)\;
%%       \If{e $>$ 0}{
%%         PushFront(QDisc, (x, e-1))
%%       }
%%       \KwRet Sample(mut(x))\;
%%     }
%%     \lElse{\KwRet Sample(gen)}
%%   }
%% \caption{\label{algo:fuzzchick:pick}CGPT Seed Selection}
%% \end{algorithm}

@@

\item \emph{Power schedule:}
%
it is not clear how the heuristic used to assign a given budget to each mutable
test case in CGPT works in the context of high-level structure-preserving
mutations.
%
If it assigns too much budget to certain not-so-interesting seeds, some bugs
might not be discovered on a timely basis.
%
Conversely, assigning too little budget to interesting test cases might mean
that some bugs cannot be discovered at all --- randomly generating the same test
case again later does not help, as it becomes uninteresting based on historic
trace information.

@@

\begin{figure}[b]
\begin{verbatim}
mutate (Leaf x)       = [ PURE (Branch def x def) ]
mutate (Branch l x r) = [ PURE l, PURE r
                        , PURE (Leaf x)
                        , PURE (Branch l x l)
                        , PURE (Branch r x r)
                        , PURE (Branch r x l) ]
\end{verbatim}
\vspace{-5pt}
\caption{\label{fig:mutagen:mutator}\mutagen mutator for the \texttt{Tree} data
  type.}
%% \vspace{-15pt}
\end{figure}

@@

\begin{verbatim}
positions (Leaf x)       = node [ (0, positions x) ]
positions (Branch l x r) = node [ (0, positions l)
                                , (1, positions x)
                                , (2, positions r) ]
\end{verbatim}

@@

\begin{equation}
  \nonumber
  \footnotesize
  \texttt{positions}\quad
  \left(
  \tikz[
    baseline=-6mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\texttt{Branch}}
      child {node {\texttt{Leaf}}
        child {node {\texttt{1}}}
      }
      child {node {\texttt{2}}}
      child {node {\texttt{Leaf}}
        child {node {\texttt{3}}}
      };
  }
  \right)
  \quad
  =
  \quad
  \tikz[
    baseline=-6mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\texttt{[]}}
      child {node {\texttt{[0]}}
        child {node {\texttt{[0,0]}}}
      }
      child {node {\texttt{[1]}}}
      child {node {\texttt{[2]}}
        child {node {\texttt{[2,0]}}}
      };
  }
\end{equation}

@@

Having the exhaustive mutation mechanism in place, we can finally introduce the
base testing loop used by \mutagen.
%
This is outlined in Algorithm \ref{algo:mutagen:loop}.
%
Like in CGPT, we use two queues, QValid and QDisc to keep valid and discarded mutable test cases,
respectively.
%
%% The testing loop enqueues and retrieves interesting test cases from them before
%% falling back to random generation.

\begin{algorithm}[t]
  \SetArgSty{textnormal}
  %% \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, R, gen, mut}}{
    i $\gets$ 0\;
    TLog, QValid, QDisc $\gets$ $\varnothing$\;
    \While{i $<$ N}{
      x $\gets$ Pick(QValid, QDisc, gen)\;
      (result, trace) $\gets$ WithTrace(P(x))\;
      \lIf{\KwNot result}{\KwRet Bug(x)}
      \If{Interesting(TLog, trace)}{
        \If{\KwNot Discarded(result)}{
          batch $\gets$ Mutate(x, mut, R)\;
          Enqueue(QValid, batch)\;
        }
        \uElseIf{\KwNot Discarded(Parent(x))}{
          batch $\gets$ Mutate(x, mut, R)\;
          Enqueue(QDisc, batch)\;
        }
      }
      i $\gets$ i+1\;
    }
    \KwRet Ok\;
  }
\caption{\label{algo:mutagen:loop}\mutagen Testing Loop}
\end{algorithm}

@@

\begin{algorithm}[t]
  \SetArgSty{textnormal}
  %% \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{Pick}{Pick}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\Pick{QValid, QDisc, gen}}{
    \If{\KwNot Empty(QValid)}{
      batch $\gets$ Deque(QValid)\;
      \If{Empty(batch)}{
        Pick(QValid, QDisc, gen)\;
      } \uElse{
        PushFront(QValid, Rest(batch))\;
        \KwRet First(batch)\;
      }
    }
    \If{\KwNot Empty(QValid)}{
      batch $\gets$ Deque(QValid)\;
      \If{Empty(batch)}{
        Pick(QValid, QDisc, gen)\;
      } \uElse{
        PushFront(QValid, Rest(batch))\;
        \KwRet First(batch)\;
      }
    }
    \lElse{\KwRet Sample(gen)}
  }
\caption{\label{algo:mutagen:pick}\mutagen Seed Selection}
\end{algorithm}

@@

\begin{algorithm}[t]
  \SetArgSty{textnormal}
  %% \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, R, gen, mut}}{
    $\cdots$\;
    x $\gets$ Pick(QValid, QDisc, gen, mut)\;
    (result, trace) $\gets$ WithTrace(P(x))\;
    $\cdots$\;
    \If{Interesting(TLog, trace)}{
      \If{\KwNot Discarded(result)}{
        batch $\gets$ Mutate(x, mut, R)\;
        {\color{red} prio $\gets$ BranchDepth(TLog, trace)}\; %% NEW
        {\color{red} PushFront(QValid, prio, batch)}\;          %% NEW
      }
      $\cdots$\;
    }
  }
  %% \vspace{5pt}
  \SetKwFunction{Pick}{Pick}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\Pick{QValid, QDisc, gen}}{
    \If{\KwNot Empty(QValid)}{
      {\color{red} (batch, prio) $\gets$ DequeMin(QValid)}\; %%NEW
      \If{Empty(batch)}{
        Pick(QValid, QDisc, gen)\;
      } \uElse{
        {\color{red} PushFront(QValid, prio, Rest(batch))}\; %% NEW
        \KwRet First(batch)\;
      }
    }
    $\cdots$\;
  }

@@

\begin{algorithm}
  \SetArgSty{textnormal}
  %% \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, gen, mut}}{
    {\color{red}
    boring  $\gets$ 0\;
    reset   $\gets$ 1000\;
    R       $\gets$ 1\;
    }
    $\cdots$\;
    \While{i $<$ N}{
      {\color{red}
      \If{boring $>$ reset}{
        TLog   $\gets$ $\varnothing$\;
        reset $\gets$ reset * 2\;
        R     $\gets$ R * 2\;
      }
      }
      $\cdots$\;
      \lIf{\KwNot result}{\KwRet Bug(x)}
      \If{Interesting(TLog, trace)}{
        {\color{red} boring  $\gets$ 0}\;
        $\cdots$\;
      }
      \lElse {
        {\color{red} boring $\gets$ boring + 1}
      }
      $\cdots$\;
    }
  }

@@

\section{Case studies}
\label{sec:casestudies}

@@

\begin{table}[t]
\scriptsize
\begin{tabular}{|c|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Category}
& \textbf{Description} \\
\hline
1
& Validator
& Bug
& Invalid memory alignment validation \\
\hline
2
& Validator
& Discrepancy
& Validator accepts blocks returning multiple values  \\
\hline
3
& Interpreter
& Bug
& Instance function invoker silently ignores arity mismatch \\
\hline
4
& Interpreter
& Bug
& Allowed out-of-bounds memory access \\
\hline
5
& Interpreter
& Discrepancy
& NaN reinterpretation does not follow reference \\
\hline
\end{tabular}
\caption{\label{table:wasm:bugs}Issues found by \mutagen in
  \textit{haskell-wasm}.}
%% \vspace{-20pt}
\end{table}

@@

In both case studies, we show how the FIFO scheduling and trace reset heuristics
described in \S \ref{sec:heuristics} affect the testing performance of
\mutagen by individually disabling them.
%
We call these variants \textit{no FIFO} and \textit{no reset}, respectively.
%
In the case of \textit{no reset}, the amount of random mutations is no longer
controlled by this heuristic, so we arbitrarily fixed it to 25 random mutations
throughout the execution of the tool.

@@

\begin{table}
\footnotesize
%% \vspace{-5pt}
\begin{tabular}{|l|c|c|c|c|}
\hline
\multicolumn{1}{|c|}{
  \multirow{2}{*}{Property}}
  & \multicolumn{2}{c|}{\quickcheck}
  & \multicolumn{2}{c|}{\mutagen} \\
  \cline{2-5}
  \multicolumn{1}{|c|}{}
  & \multicolumn{1}{c|}{total}
  & passed
  & \multicolumn{1}{c|}{total}
  & passed \\
\hline
\textit{prop\_validator}
& 41374.92
& 0.00029
& 2243.26
& 519.83\\
\hline
\textit{prop\_interpreter\_i32}
& 134000
& 23.16
& 1815.81
& 351.16 \\
\hline
\end{tabular}
\caption{\label{table:wasm:overhead}Mean number of tests per second executed by
  each tool on the WebAssembly case study.}
%% \vspace{-10pt}
\end{table}

@@

In this work we use exhaustiveness as a way to reliably enforce that all
possible mutants of an interesting seed are executed.
%
In contrast to fully exhaustive testing tools, \mutagen relies on randomness to
find initial interesting mutable test cases.
%
In terms of pruning, it is possible to instruct \mutagen to detect unused
subexpressions (using a technique similar to that of \emph{LazySmallCheck}) to
avoid scheduling mutations over their corresponding positions.
%
This could, in principle, improve the overall performance when testing
properties that tend to short-circuit, leaving parts of the input unevaluated.
%
In our case studies, however, we observed that their preconditions tend to be
quite strict when executing test cases obtained by mutating valid existing ones,
fully evaluating their input before executing the postcondition.
%
Thus, we do not forsee considerable improvements by applying this idea in this
particular scenario.
%
Gathering empirical evidence about using pruning via lazy evaluation in \mutagen
is an effort that we keep as future work.

@@

As indicated throughout this work, there are several directions for future work:
%
combining \mutagen with specialized generators; improving the code
instrumentation in order to target mutations towards specific parts of the code;
and evaluating the effect of lazy pruning; among others.
%
In addition, we want to investigate how to redefine our automatically
synthesized mutators in a stateful manner.
%
This way, it would be possible to apply mutations that preserve complex
properties of the generated data simply by construction, e.g., mutations that
\emph{always} produce well-typed programs where variables are always in scope.
%
The main challenge will be to achieve this while keeping the testing process as
automatable as possible.

>>>>

\usepackage{wrapfig}
\usepackage{xcolor}
\usepackage{color}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{caption}
\usepackage[ruled,vlined,noend]{algorithm2e}
\usepackage{xspace}
\usepackage{tikz}
\usepackage{tikzscale}
\usepackage{pgfplots}
\usepackage{paralist}
\usepackage{enumitem}
\usepackage{titlesec}

\titlespacing*{\section}{0pt}{5pt}{5pt}
\titlespacing*{\subsection}{0pt}{5pt}{5pt}
\titlespacing*{\subsubsection}{0pt}{5pt}{5pt}
\titlespacing*{\paragraph}{10pt}{0pt}{10pt}

@@

\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}

@@

% gen_tree :: Size -> Gen a -> Gen (Tree a)

\begin{verbatim}
genTree 0 = Leaf <$> genA
genTree n = oneof
  [Leaf <$> genA, Branch <$> genTree(n-1) <*> genA <*> genTree(n-1)]
\end{verbatim}

\noindent This definition follows a common type-directed generation approach,
and it is a good example of what to expect from a random generator synthesized
automatically from the codebase.
%
In esence, this generator randomly picks one data constructor from its
corresponding data type definition with uniform probability (i.e, either a
\texttt{Leaf} or a \texttt{Branch}), and then calls itself for every recursive
subexpression needed to produce a well-typed value, carefully reducing the size
limit by a unit on each step, ensuring termination by restricting itself to
generating leaves when it reaches zero.
%
Payloads are generated simply by delegating the task to an external random
generator (\texttt{genA}) defined elsewhere.
%

@@

Readers familiar with RBPT will recognize that using \texttt{genTree} to test
\texttt{prop\_insert\_inv} with \quickcheck directly does not work well.
%
In practice, \quickcheck gives up if most of the generated test cases get
discarded due to failed preconditions.
%
%because they do not satisfy the balanced precondition.
%
Sadly, most of the inputs produced by our na\"ive generator suffer from this
problem, and the interesting part of the property (\texttt{balanced (insert x
  t)}) is tested very sporadically as a result.
%% %
%% In other words, \emph{the precondition of our testing property becomes very
%%   sparse when we use simple, automatically derived generators.}

@@

%% Once inside of the main loop, the algorithm picks the next test case using a
%% simple criterion: if there are valid values enqueued for mutation, it picks the
%% first one, mutates it and returns it, decreasing its energy by one.
%% %
%% If \textit{QValid} is empty, then the same is attempted using \textit{QDiscarded}.
%% %
%% If none of the mutation queues contain any candidates, CGPT generates a new
%% value from scratch.
%% %
%% This selection process is illustrated in detail in Algorithm
%% \ref{algo:fuzzchick:pick}.

@@

%% \paragraph{Testing Loop}
%% Outlined in Algorithm \ref{algo:fuzzchick:loop}, CGPT starts by creating two
%% queues, \textit{QValid} and \textit{QDiscarded} for valid and discarded previously
%% executed test cases, respectively.
%% %
%% Enqueued values are stored along with a given mutation budget, that controls how
%% many times a given test case can be mutated before being finally thrown away.

%% Once inside of the main loop, the algorithm picks the next test case using a
%% simple criterion: if there are valid values enqueued for mutation, it picks the
%% first one, mutates it and returns it, decreasing its energy by one.
%% %
%% If \textit{QValid} is empty, then the same is attempted using \textit{QDiscarded}.
%% %
%% If none of the mutation queues contain any candidates, CGPT generates a new
%% value from scratch.
%% %
%% This selection process is illustrated in detail in Algorithm
%% \ref{algo:fuzzchick:pick}.

@@

%% \begin{algorithm}
%%   \SetArgSty{textnormal}
%%   %% \SetInd{0em}{0.75em}
%%   \SetKw{KwNot}{not}
%%   \SetAlgoLined
%%   \DontPrintSemicolon
%%   \SetKwFunction{FC}{Loop}
%%   \SetKwProg{Fn}{Function}{:}{}
%%   \Fn{\FC{P, N, gen, mut}}{
%%     i $\gets$ 0\;
%%     TLog, QValid, QDiscarded $\gets$ $\varnothing$\;
%%     \While{i $<$ N}{
%%       x $\gets$ Pick(QValid, QDiscarded, gen, mut)\;
%%       (result, trace) $\gets$ WithTrace(P(x))\;
%%       \lIf{\KwNot result}{\KwRet Bug(x)}
%%       \If{Interesting(TLog, trace)}{
%%         e $\gets$ Energy(TLog, x, trace)\;
%%         \eIf{\KwNot Discarded(result)}{
%%           Enqueue(QValid, (x, e))\;
%%         }{
%%           Enqueue(QDiscarded, (x, e))\;
%%         }
%%       }
%%       i $\gets$ i+1\;
%%     }
%%     \KwRet Ok\;
%%   }
%% \caption{\label{algo:fuzzchick:loop}CGPT Testing Loop}
%% \end{algorithm}

%% \begin{algorithm}
%%   \SetArgSty{textnormal}
%%   %% \SetInd{0em}{0.75em}
%%   \SetKw{KwNot}{not}
%%   \SetAlgoLined
%%   \DontPrintSemicolon
%%   \SetKwFunction{Pick}{Pick}
%%   \SetKwProg{Fn}{Function}{:}{}
%%   \Fn{\Pick{QValid, QDiscarded, gen, mut}}{
%%     \If{\KwNot Empty(QValid)}{
%%       (x,e) $\gets$ Deque(QValid)\;
%%       \If{e $>$ 0}{
%%         PushFront(QValid, (x, e-1))
%%       }
%%       \KwRet Sample(mut(x))\;
%%     }
%%     \uElseIf{\KwNot Empty(QDiscarded)}{
%%       (x,e) $\gets$ Deque(QDiscarded)\;
%%       \If{e $>$ 0}{
%%         PushFront(QDiscarded, (x, e-1))
%%       }
%%       \KwRet Sample(mut(x))\;
%%     }
%%     \lElse{\KwRet Sample(gen)}
%%   }
%% \caption{\label{algo:fuzzchick:pick}CGPT Seed Selection}
%% \end{algorithm}

@@

\item \emph{Power schedule:}
%
it is not clear how the heuristic used to assign a given budget to each mutable
test case in CGPT works in the context of high-level structure-preserving
mutations.
%
If it assigns too much budget to certain not-so-interesting test cases, some
bugs might not be discovered on a timely basis.
%
Conversely, assigning too little budget to interesting test cases might mean
that some bugs cannot be discovered at all --- randomly generating the same test
case again later does not help, as it becomes uninteresting based on historic
trace information.

@@

\begin{figure}[t]
\begin{verbatim}
mutate (Leaf x) =
  [ PURE (Branch def x def) ]
mutate (Branch l x r) =
  [ PURE l, PURE r
  , PURE (Leaf x)
  , PURE (Branch l x l), PURE (Branch r x r), PURE (Branch r x l) ]
\end{verbatim}
\vspace{-5pt}
\caption{\label{fig:mutagen:mutator}\mutagen mutator for the \texttt{Tree} data
  type.}
\vspace{-5pt}
\end{figure}

@@

\begin{verbatim}
positions (Leaf x) =
  node [ (0, positions x) ]
positions (Branch l x r) =
  node [ (0, positions l), (1, positions x), (2, positions r) ]
\end{verbatim}

@@

\vspace{-5pt}
\begin{equation}
  \nonumber
  \footnotesize
  \texttt{positions}\quad
  \left(
  \tikz[
    baseline=-6mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\texttt{Branch}}
      child {node {\texttt{Leaf}}
        child {node {\texttt{1}}}
      }
      child {node {\texttt{2}}}
      child {node {\texttt{Leaf}}
        child {node {\texttt{3}}}
      };
  }
  \right)
  \quad
  =
  \quad
  \tikz[
    baseline=-6mm,
    level distance=0.5cm,
    level 1/.style={sibling distance=0.5cm},
    level 2/.style={sibling distance=0.5cm}
  ]{
    \node {\texttt{[]}}
      child {node {\texttt{[0]}}
        child {node {\texttt{[0,0]}}}
      }
      child {node {\texttt{[1]}}}
      child {node {\texttt{[2]}}
        child {node {\texttt{[2,0]}}}
      };
  }
\end{equation}

@@

Having the exhaustive mutation mechanism in place, we can finally introduce the
base testing loop used by \mutagen.
%
This is outlined in Algorithm \ref{algo:mutagen:loop}.
%
Like in CGPT, we use two queues, QValid and QDiscarded to keep valid and discarded mutable test cases,
respectively.
%
%% The testing loop enqueues and retrieves interesting test cases from them before
%% falling back to random generation.

\begin{algorithm}[t]
  \SetArgSty{textnormal}
  %% \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, R, gen, mut}}{
    i $\gets$ 0\;
    TLog, QValid, QDiscarded $\gets$ $\varnothing$\;
    \While{i $<$ N}{
      x $\gets$ Pick(QValid, QDiscarded, gen)\;
      (result, trace) $\gets$ WithTrace(P(x))\;
      \lIf{\KwNot result}{\KwRet Bug(x)}
      \If{Interesting(TLog, trace)}{
        \If{\KwNot Discarded(result)}{
          batch $\gets$ Mutate(x, mut, R)\;
          Enqueue(QValid, batch)\;
        }
        \uElseIf{\KwNot Discarded(Parent(x))}{
          batch $\gets$ Mutate(x, mut, R)\;
          Enqueue(QDiscarded, batch)\;
        }
      }
      i $\gets$ i+1\;
    }
    \KwRet Ok\;
  }
\caption{\label{algo:mutagen:loop}\mutagen Testing Loop}
\end{algorithm}

@@

\begin{algorithm}[t]
  \SetArgSty{textnormal}
  %% \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{Pick}{Pick}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\Pick{QValid, QDiscarded, gen}}{
    \If{\KwNot Empty(QValid)}{
      batch $\gets$ Deque(QValid)\;
      \If{Empty(batch)}{
        Pick(QValid, QDiscarded, gen)\;
      } \uElse{
        PushFront(QValid, Rest(batch))\;
        \KwRet First(batch)\;
      }
    }
    \If{\KwNot Empty(QValid)}{
      batch $\gets$ Deque(QValid)\;
      \If{Empty(batch)}{
        Pick(QValid, QDiscarded, gen)\;
      } \uElse{
        PushFront(QValid, Rest(batch))\;
        \KwRet First(batch)\;
      }
    }
    \lElse{\KwRet Sample(gen)}
  }
\caption{\label{algo:mutagen:pick}\mutagen Seed Selection}
\end{algorithm}

@@

\begin{algorithm}[t]
  \SetArgSty{textnormal}
  %% \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, R, gen, mut}}{
    $\cdots$\;
    x $\gets$ Pick(QValid, QDiscarded, gen, mut)\;
    (result, trace) $\gets$ WithTrace(P(x))\;
    $\cdots$\;
    \If{Interesting(TLog, trace)}{
      \If{\KwNot Discarded(result)}{
        batch $\gets$ Mutate(x, mut, R)\;
        {\color{red} prio $\gets$ BranchDepth(TLog, trace)}\; %% NEW
        {\color{red} PushFront(QValid, prio, batch)}\;          %% NEW
      }
      $\cdots$\;
    }
  }
  %% \vspace{5pt}
  \SetKwFunction{Pick}{Pick}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\Pick{QValid, QDiscarded, gen}}{
    \If{\KwNot Empty(QValid)}{
      {\color{red} (batch, prio) $\gets$ DequeMin(QValid)}\; %%NEW
      \If{Empty(batch)}{
        Pick(QValid, QDiscarded, gen)\;
      } \uElse{
        {\color{red} PushFront(QValid, prio, Rest(batch))}\; %% NEW
        \KwRet First(batch)\;
      }
    }
    $\cdots$\;
  }

@@

\begin{algorithm}[t]
  \SetArgSty{textnormal}
  %% \SetInd{0em}{0.75em}
  \SetKw{KwNot}{not}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwFunction{MG}{Loop}
  \SetKwProg{Fn}{Function}{:}{}
  \Fn{\MG{P, N, gen, mut}}{
    {\color{red}
    boring  $\gets$ 0\;
    reset   $\gets$ 1000\;
    R       $\gets$ 1\;
    }
    $\cdots$\;
    \While{i $<$ N}{
      {\color{red}
      \If{boring $>$ reset}{
        TLog   $\gets$ $\varnothing$\;
        reset $\gets$ reset * 2\;
        R     $\gets$ R * 2\;
      }
      }
      $\cdots$\;
      \lIf{\KwNot result}{\KwRet Bug(x)}
      \If{Interesting(TLog, trace)}{
        {\color{red} boring  $\gets$ 0}\;
        $\cdots$\;
      }
      \lElse {
        {\color{red} boring $\gets$ boring + 1}
      }
      $\cdots$\;
    }
  }

@@

%% \vspace{-10pt}
\section{Case studies}
\label{sec:casestudies}

@@

\begin{table}[t]
\scriptsize
\begin{tabular}{|c|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Category}
& \textbf{Description} \\
\hline
1
& Validator
& Bug
& Invalid memory alignment validation \\
\hline
2
& Validator
& Discrepancy
& Validator accepts blocks returning multiple values  \\
\hline
3
& Interpreter
& Bug
& Instance function invoker silently ignores arity mismatch \\
\hline
4
& Interpreter
& Bug
& Allowed out-of-bounds memory access \\
\hline
5
& Interpreter
& Discrepancy
& NaN reinterpretation does not follow reference \\
\hline
\end{tabular}
\caption{\label{table:wasm:bugs}Issues found by \mutagen in
  \textit{haskell-wasm}.}
\vspace{-10pt}
\end{table}

@@

In both case studies, we show how the two \mutagen heuristics described in \S
\ref{sec:heuristics} affect the testing performance by individually disabling
them.
%
We call these variants \textit{no FIFO} and \textit{no reset}, respectively.
%
In the case of \textit{no reset}, the amount of random mutations is no longer
controlled by this heuristic, so we arbitrarily fixed it to 25 random mutations
throughout the testing loop.

@@

\begin{table}
\footnotesize
%% \vspace{-5pt}
\begin{tabular}{|l|c|c|c|c|}
\hline
\multicolumn{1}{|c|}{
  \multirow{2}{*}{Property}}
  & \multicolumn{2}{c|}{\quickcheck}
  & \multicolumn{2}{c|}{\mutagen} \\
  \cline{2-5}
  \multicolumn{1}{|c|}{}
  & \multicolumn{1}{c|}{total}
  & passed
  & \multicolumn{1}{c|}{total}
  & passed \\
\hline
\textit{prop\_validator}
& 41374.92
& 0.00029
& 2243.26
& 519.83\\
\hline
\textit{prop\_interpreter\_i32}
& 134000
& 23.16
& 1815.81
& 351.16 \\
\hline
\end{tabular}
\caption{\label{table:wasm:overhead}Mean number of tests per second executed by
  each tool on the WebAssembly case study.}
\vspace{-10pt}
\end{table}

@@

In this work we use exhaustiveness as a way to reliably enforce that all
possible mutants of an interesting test case are executed.
%
In contrast to fully exhaustive testing tools, \mutagen relies on randomness to
find initial interesting mutable test cases.
%
In terms of pruning, it is possible to instruct \mutagen to detect unused
subexpressions (using a technique similar to that of \emph{LazySmallCheck}) to
avoid scheduling mutations over their corresponding positions.
%
This could, in principle, improve the overall performance when testing
properties that tend to short-circuit, leaving parts of the input unevaluated.
%
In our case studies, however, we observed that their preconditions tend to be
quite strict when executing test cases obtained by mutating valid existing ones,
fully evaluating their input before executing the postcondition.
%
Thus, we do not forsee considerable improvements by applying this idea in this
particular scenario.
%
Gathering empirical evidence about using pruning via lazy evaluation in \mutagen
is an effort that we keep as future work.

@@

As indicated throughout this work, there are several directions for future work:
%
combining \mutagen with specialized generators; improving the code
instrumentation in order to target mutations towards specific parts of the code;
and evaluating the effect of lazy pruning; among others.
%% %
%% In addition, we want to investigate how to redefine our automatically
%% synthesized mutators in a stateful manner.
%% %
%% This way, it would be possible to apply mutations that preserve complex
%% properties of the generated data simply by construction, e.g., mutations that
%% \emph{always} produce well-typed programs where variables are always in scope.
%% %
%% The main challenge will be to achieve this while keeping the testing process as
%% automatable as possible.
